\section{Does PTS Fulfill The Requirements of PT?}\label{sec:does-pts-fulfill-the-requirements-of-pt?}

This thesis is concerned about implementing Package Templates in TypeScript.
However, in order to determine to what degree we have actually implemented PT or just created something that looks like it, we have to understand what the requirements of PT are, and if we are meeting those requirements.
We will therefore in this chapter look at the requirements as described in~\cite{jot}.
After getting an understanding of the requirements we are going to look at how our implementation holds up to them.

\input{chapters/discussion/the-requirements-of-pt.tex}

\subsection{PTS' Implementation of the Requirements}\label{subsec:pts'-implementation-of-the-requirements}

With a proper understanding of the requirements of PT we can examine our implementation and see whether our implementation fulfills these requirements.
For each requirement we will be looking at a program in PTS which showcases the requirements, and the resulting program after compilation.


\subsubsection{Parallel Extension}\label{subsubsec:pts-parallel-extension}

Fulfilling the requirement of parallel extension requires us to be able to merge or make extensions to classes, and be able to reference these additions within the same package/template, in the same or in a different class.
To understand how this requirement can be fulfilled it is important to understand how the requirement could fail to be fulfilled.
A failure to fulfill the requirement would be that making additions in parallel would fail to compile, or create an otherwise incorrect program.
Failure to compile might of course not always be a bad thing, there are certain scenarios where we do want the compiler to throw an error.
There are mainly two scenarios where we would like the compilation to fail for additions:
\begin{itemize}
    \item trying to make an addition to a non-existent class, and
    \item trying to reference non-existent attributes in a class.
\end{itemize}

The first scenario where our compiler should fail is when we are making additions to a non-existent class.
This will be caught in the class merging part of our compiler.
In the class merging part of the implementation the compiler will group all class declarations and additions by the class name.
If there is a group containing only additions then it will fail, as we have no class to make additions to.

The second scenario is when we are trying to reference non-existent attributes in a class.
An example of this can be seen in listing~\vref{lst:pts-parallel-extension-non-existent-attribute}.
This example will fail during the type-checking of our packages/templates, as discussed in~\vref{sec:type-checking-of-templates}.
Our approach for dealing with this is pretty much by not dealing with it, and instead assume that everything is okay at this stage of the compilation.
We will then instead discover any inconsistencies in the type-checking stage of the compiler.
In the aforementioned listing it is of course pretty easy to examine class \codeword{A} to see if it contains an attribute \codeword{h}, however it might not always be this easy.
In a more complicated example where we are in the process of merging several classes and additions it might prove a tougher task to see if the addition would result in a type-safe class.
So as long as we are able to perform the addition we can assume that it is working as intended and instead let the TypeScript compiler check if it is type-safe, after the addition has been performed.

\begin{code}{typescript}{An example showing a program that should fail during compilation, where we are trying to reference a non-existent attribute, \codeword{h}, in an addition to class \codeword{A}.}{lst:pts-parallel-extension-non-existent-attribute}
    template T {
        class A {
            function f() {
                return 1;
            }
        }
    }

    package P {
        inst T;
        addto A {
            function g() {
                this.h();
            }
        }
    }
\end{code}

Now that we understand when we want compilation to fail let us look at where we do not want it to fail, when we have a valid parallel extension.
One way it could result in an unwanted failure is if we tried to check if the addition contains any invalid references or type errors.
This could commonly happen if we are trying to check the addition's references to the declared class.
However, as discussed above, checking if a reference to an attribute is valid is quite tricky, and in our implementation instead leaves this up to the TypeScript compiler.
By doing this we will not incorrectly throw any false-negatives when it comes to parallel extensions.
This approach does unfortunately come with some downsides.
By not addressing the issue at the addition stage it makes it harder to give informative error messages when invalid references do occur, however this was a tradeoff that was beneficial for this project.

\subsubsection{Hierarchy Preservation}\label{subsubsec:pts-hierarchy-preservation}

In order to fulfill the hierarchy preservation requirement we have to preserve all super-/subclass relations after additions and merges have been applied.
Listing~\vref{lst:pts-hierarchy-preservation} shows a program, and the resulting TypeScript program after compilation, which fulfills the requirement of hierarchy preservation.
This one example fulfills the requirement as class \codeword{B} is still a subclass to class \codeword{A} after both a merge and an addition is made to \codeword{B}.
As we talked about briefly in section~\vref{subsec:merging-class-declarations} when we merge classes we make sure to also merge their class heritage, combining the extending classes and implementing interfaces of the different classes.
This means that we might end up with instances where we are extending multiple different classes, however this will then be picked up in the type-checking stage of the compiler.
If we had not merged class heritage, we could have ended up breaking the inheritance hierarchy in the aforementioned listing, as we could have for example ended up with class \codeword{C}'s heritage, which does not have a superclass.
Because of the heritage merging we can with confidence say that we have fulfilled the requirement of hierarchy preservation, as we always preserve all super-/subclass relations.

\begin{code}{typescript}{Example showcasing a program where the super-/subclass relation between classes \codeword{A} and \codeword{B} is preserved after additions and class merging have been applied. We can see the resulting TypeScript program at the bottom of the listing, where the semantics are as expected.}{lst:pts-hierarchy-preservation}
    // PTS
    template T1 {
        class A {
            i = 0;
        }

        class B extends A {
            f() {
                return this.i;
            }
        }
    }

    template T2 {
        class C {
            j = 0;
        }
    }

    package P {
        inst T1;
        inst T2 { C -> B };
        addto B {
            k = 0;
        }
    }

    // Resulting program
    class A {
        i = 0;
    }

    class B extends A {
        f() {
            return this.i;
        }
        j = 0;
        k = 0;
    }
\end{code}

\subsubsection{Renaming}\label{subsubsec:pts-renaming}

In order to be able to fulfill the renaming requirement our implementation should be able to rename classes and their attributes.
This renaming should result in a program where not only the declarations have been renamed, but also all references.
Listing~\vref{lst:pts-renaming} shows a program with an example of renaming in PTS, where we are renaming a class, \codeword{A}, and the class' attribute, \codeword{i}.
We can see in the resulting program that the identifier in the declaration of both the class and the attribute has changed, as well ts he references to these in the constructor of the class and references in another class, \codeword{B}\@.
The renaming has also not wrongly renamed other references that are similar in naming, such as the parameter of the constructor of class \codeword{A}\@.

\begin{code}{typescript}{Example of renaming in PTS.}{lst:pts-renaming}
    // PTS
    template T {
        class A {
            i = 0;
            constructor(i: number) {
                this.i = i;
            }
        }

        class B {
            a = new A();
            function f() {
                return a.i;
            }
        }
    }

    pack P {
        inst T { A -> X (i -> j) };
    }

    // Resulting program
    class X {
        j = 0;
        constructor(i: number) {
            this.j = i;
        }
    }

    class B {
        a = new X();
        function f() {
            return a.j;
        }
    }
\end{code}

Because of TypeScript's structural type system, renaming will have to work a bit differently than it would in a nominally typed language.
In nominal PT, all references are always renamed.
In structural PT we could end up with programs where some references can not be renamed, as some variables might not be directly tied with a class.
Listing~\vref{lst:pts-renaming-problem} showcases this problem.
The problem arises in the \codeword{a} attribute of class \codeword{B} in template \codeword{T}.
This has been declared to be a variable expecting an object where there exists an attribute \codeword{i}.
\codeword{a} is initialised with an object of the class \codeword{A}.
This is fine in template \codeword{T}, however when \codeword{T} is instantiated in package \codeword{P}, and \codeword{A}'s attribute \codeword{i} is renamed to \codeword{j}, this is no longer the case.
The \codeword{a} variable in class \codeword{B} is not a direct reference to class \codeword{A}, and we can thus not rename the reference to the attribute \codeword{i}.
Since an object of \codeword{A} no longer contains an attribute \codeword{i}, this is no longer a valid value for variable \codeword{a}.



\begin{code}{typescript}{Example showcasing the problem of having renaming in a structural language. In class \codeword{B} we have an attribute, \codeword{a}, that expects an object that contains an attribute \codeword{i}. The attribute is initialized with an \codeword{A} object. This is fine in template \codeword{T} as \codeword{A} contains an attribute \codeword{i}, however when class \codeword{A}'s attribute is renamed in the instantiation in package \codeword{P} then an object of \codeword{A} is no longer valid as a value, since it no longer contains an attribute \codeword{i}. This is an instance where we can't just rename the references to \codeword{i}, since this reference isn't explicitly related to \codeword{A}.}{lst:pts-renaming-problem}
    // PTS
    template T {
        class A {
            i = 0;
        }

        class B {
            a : { i : number } = new A();
            i = a.i;
        }
    }

    pack P {
        inst T { A -> A (i -> j) };
    }

    // Expected result
    class A {
        j = 0;
    }

    class B {
        a : { i : number } = new A();
        i = a.i;
    }
\end{code}

The aforementioned listing shows how we would like the compilation result to look like, however this is not the result the current implementation produces.
TypeScript's type system can be quite complicated, and due to a lack of time I chose to ignore most of the type declarations.
The current implementation would have treated the attribute \codeword{a} as an \codeword{A}-variable, since it is being initialized with an object of \codeword{A}, and therefore have renamed later references to \codeword{a.i} to \codeword{a.j}.
It was more important to get a working prototype, than support all scenarios with different type signatures.
This is something I would of course have liked to take into consideration if I had more time to spend on the implementation.
Deciding the type of variables is something that possibly would have come for cheaper if I had opted for a fork of the TypeScript compiler as my approach.
This is something we will come back to in~\vref{subsec:result-approach}.

Nominal PT however is also not perfect in this regard, as we can run into issues with renamings of interfaces' attributes.
If a class implements an interface, and we in a later instantiation rename the interfaces' attributes, and not the attributes of the implementing class, we could end up with a program where the class no longer fulfills the requirement of the interface.
In listing~\vref{lst:nominal-pt-interface-problem} we can see this problem in action, where after a rename of interface \codeword{I}'s method \codeword{f}, the class \codeword{A} no longer implements the interface properly.
This is of course also a problem in structural PT, and can potentially be an even bigger problem as we would not necessarily have an explicit relation between the interface and the class, and could therefore also not properly describe the error to the programmer.

\begin{code}{java}{Example showing how a renaming of an interfaces' attributes could result in an invalid program.}{lst:nominal-pt-interface-problem}
    template T {
        interface I {
            void f();
        }

        class A implements I {
            void f() { ... }
        }
    }

    package P {
        inst T with I => I (f() -> g());
        // Error: A is not abstract and does not
        // override abstract method f() in I
    }
\end{code}

\subsubsection{Multiple Uses}\label{subsubsec:pts-multiple-uses}

In order for this requirement to be fulfilled we should be able to re-use a template several times, with different renamings and additions while the different instantiations stay independent of each other.
This was something I paid extra attention to during implementation, not just to fulfill the requirement, but to avoid bugs.
I solved this by making sure that while transforming the AST this would be done in an immutable fashion.
In order to test this we will be creating a simple program where we instantiate the same template more than once and see if the resulting program is as expected.
The program can be seen in listing~\vref{lst:pts-multiple-uses}.
The program comprises a template \codeword{T} with a single class, \codeword{A}, with an attribute \codeword{i}.
This template will then be instantiated three times, where we first will be renaming the class and field, then instantiate without renaming, and finally instantiate it with just an attribute renaming.
The expected program should have two classes, one class \codeword{B}, with an attribute \codeword{j}, and a class \codeword{A} where the two bottom instantiations should have created a merged class with attributes \codeword{i} and \codeword{x}.
We can see from the resulting program after a successful compilation that this is as expected.

\begin{code}{typescript}{A program showcasing multiple uses in PTS, and the resulting program in TypeScript at the bottom.}{lst:pts-multiple-uses}
    // PTS
    template T {
        class A {
            i = 0;
        }
    }

    pack P {
        inst T { A -> B (i -> j) };
        inst T;
        inst T { A -> A (i -> x) };
    }

    // Resulting program
    class B {
        j = 0;
    }

    class A {
        i = 0;
        x = 0;
    }
\end{code}

\subsubsection{Type Parameterization}\label{subsubsec:pts-type-parameterization}

The type parameterization requirement is something the implementation does not fulfill.
This was not implemented due to it not being prioritized.
There is only so much time available during the span of a master thesis, and I chose to look at how the core of PT would fit into a structurally typed language like TypeScript, rather than on making sure it would be a fully fleshed out implementation of PT\@.
Another reason for avoiding this is that much of type parameterization can be achieved through merging and renaming.
Listing~\vref{lst:type-parameterization-without-rt} shows an example of how you can use an empty class as a generic type implementation of lists, similar to the list implementation with required types in listing~\vref{lst:type-parameterization}.
Required types do of course have a lot of advantages such as making it possible to constrain the type, and forcing the programmer to give an actual parameter for the type, which we are unable to do.

\begin{code}{java}{Example of a similar list implementation as in listing~\vref{lst:type-parameterization}, without the use of required types. Instead of giving a type for the required type we will have to merge the class \codeword{E} with the "actual parameter".}{lst:type-parameterization-without-rt}
    template ListsOf {
        class E { }
        class List {
            AuxElem first, last;
            void insertAsLast(E e) { ... }
            E removeFirst() { ... }
        }
        class AuxElem {
            AuxElem next;
            E e;
        }
    }
\end{code}

\subsubsection{Class Merging}\label{subsubsec:pts-class-merging}

%TODO Les over og finpuss på dette

In order to fulfill the requirement of class merging we will have to be able to merge classes through instantiating two or more templates containing a class with a common name, or through renaming a class at instantiating leading to two classes having the same name.
A failure to fulfill this requirement could be that a resulting program would contain two different class declarations, where these should have been merged.
This could occur if we had for instance not respected the renaming of a class at the time class merging, that is if we had renamed a class to be merged, but this was not picked up during class merging.
Our implementation of class merging has been made so that it is essentially unaware of the instantiation and renaming steps, it is done as a separate step after these actions have been performed.
What this means is that after performing all instantiations, with optional renamings, we can momentarily end up with several class declarations for the same class, which would be an invalid TypeScript program, however before the closing of templates step is done we will merge the class declarations and \codeword{addto}-statements representing the same class, forming new classes.
The resulting program could of course still be an invalid TypeScript program, as the merge of classes could lead to erroneous code, such as duplicate function declarations, however this is not a result of an invalid implementation, but rather an error as a product of an invalid input program.
The class merging step, will merge classes without checking their bodies, as these will be picked up in the type-checking step later on.
By separating the logic of instantiations and renaming, as we have done, we should minimize the potential for an invalid implementation of class merging.

Listing~\vref{lst:req-class-merging-example} shows a program with class merging, and the resulting program from a compilation of the PTS program.
Here we merge two classes with the same name, \codeword{A}, which has two different

\begin{code}{typescript}{Example of a program with class merging, where we merge two classes, \codeword{A}, with attributes, \codeword{i} and \codeword{j}, respectively. The resulting program shown at the bottom is as expected a merge of the two classes.}{lst:req-class-merging-example}
    // PTS
    template T {
        class A {
            i = 0;
        }
    }

    pack P {
        inst T;
        inst T { A -> A (i -> j) };
    }

    // Resulting program
    class A {
        i = 0;
        j = 0;
    }
\end{code}

\subsubsection{Collection-level Type-checking}\label{subsubsec:implementation-collection-level-type-checking}

To fulfill this requirement our implementation just needs to perform a type-check on each package/template individually.
In the type-checking step of our compiler we do just this, and we therefore fulfill this requirement.

However, as we previously touched upon in section~\vref{subsec:why-typescript-verifying-templates}, TypeScript's type system is unsound.
What this means is that TypeScript's type-system will not be able to pick up all type errors.
With an unsound type-check, can we truly say we have fulfilled the requirement of collection-level type-checking?

In the proposal of PT, Java is used as the host language.
Java has also been shown to have an unsound type system~\cite{java-unsound}, so we could argue that if the requirement is fulfilled in PT, then our unsound type-check should also fulfill the requirement.

\subsection{Conclusion}\label{subsec:requirements-conclusion}

While we do not fulfill every requirement, we do fulfill most of them.
The current implementation might not be a full implementation of PT, but we can confidently say we have at least made an implementation of the core of PT for TypeScript.
Not having a full implementation does mean that we might not be able to examine all the differences between our implementation and PT, however we will be able to examine the common elements, which covers the most interesting parts.
This allows us to explore how a mechanism like PT fits with the TypeScript language, and its potential utility.